---
title: "Effect Size Transformation"
author: 
  - name: "Tobiasz Tranwinski"
    affiliation : "1"
    email: "tobiasz.tranwinski@nyu.edu"
  - name: "[Haiyang Jin](https://haiyangjin.github.io/)"
    affiliation: "1"
    address: "New York University Abu Dhabi, Saadiyat Isalnd, Abu Dhabi, United Arab Emirates"
    email: "haiyang.jin@nyu.edu"
affiliation:
  - id            : "1"
    institution   : "New York Univesity Abu Dhabi"
date: "`r format(Sys.time(), '%d-%m-%Y')`"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: true
version: "1.0"
---

<style>
pre {
overflow-x: auto;
}
pre code {
word-wrap: normal;
white-space: pre;
}
</style>

We are currently working on the meta-analyze (link). The effect size of our main interest is the product of the interaction between two or three conditions. Unfortunately, many older papers do not report the effect size. 

In this post, we discuss (1)  how to obtain the effect size of the interaction between two conditions and (2) propose the transformation of the partial eta squared to Cohen's d and then to Hedges' g.


```{r include=FALSE}
# load library
library(tidyverse)
library(afex)
```

First, we simulated the database for the 2 X 2 within-subject design:

```{r}
# simulate data
set.seed(42)
nSubj <- 20
Subjects <-factor(rep(1:nSubj,each=4))  

DV<-c(rnorm(nSubj, mean = 15, sd = 2), rnorm(nSubj, mean = 10, sd = 2))
A <-rep(c("level1","level2"),each=2,20)
B <-rep(c("level1","level2"),each=1,40)

simu_long <-data.frame(Subjects,DV,A,B)
head(simu_long)
```

The database has twenty participants. Each participant completed both conditions A (2: level1 vs. level2) and B (2: level1 vs. level2). There is eighty data points in total.

# Analyzing Data: ANOVA

To analyze data with a 2 X 2 within-subject design, we usually use ANOVA. To do so, we used package 'afex' (Singmann et al., 2021):
. 

```{r}
ANOVAresults <-aov_4(DV ~ A*B+(A*B|Subjects), data = simu_long)
nice(ANOVAresults, es = NULL, sig_symbols = rep("", 4))
```


# Effect size: partial eta-squared.  

One of the most common measures of the effect size in psychology is partial eta squared. The partial eta squared formula is relatively simple:

$$
\begin{aligned}
\eta_{p}^2 &= \frac{\mathrm{SS}_\mathrm{effect}}{\mathrm{SS}_\mathrm{effect}+\mathrm{SS}_\mathrm{error}}\\  
\end{aligned}
$$

In many cases, psychological papers do not report SS-values, we also do not have access to the actual data. Fortunately, we can calculate partial eta squared from F-value and degrees of freedom (Lakens, 2013). 

$$
\begin{aligned}
\eta_{p}^2 &= \frac{\mathrm{F*df}_\mathrm{num}}{\mathrm{F*df}_\mathrm{num}+\mathrm{df}_\mathrm{den}}\\
\mathrm{df}_\mathrm{num} &= \mathrm{(k_A-1)*(k_B-1)}\\
\mathrm{df}_\mathrm{den} &= \mathrm{N-(k_A*k_B)}\\
\end{aligned}
$$

where F is the F-value, k_A and K_B reflects number of levels in condition A (i.e., 2) and in condition B (i.e., 2), N is the number of subjects (i.e., 20).

Following this equation, partial eta squared for the interaction between condition A and B is: 

```{r}
n2p = 13.02*1/(13.02*1+19)
n2p
```

which is the same as partial eta squared, which we have already gotten: 

```{r}
nice(ANOVAresults, es = 'pes', sig_symbols = rep("", 4))
```


# Effect size transformation 

To conduct meta-analyses, it is necessary to combine effect sizes from different studies that used different metrics. To do so, we may normalize the effect sizes by transforming them into Hedges' g (Hedges & Olkin, 1985).

## Converting partial eta-squared to eta-squared

First, we can try to transform partial eta squared to eta squared. Next transform eta squared into Cohen's d and finally, transform Cohen's d to Hedges' g using the 'esc' R package (Lüdecke, 2019). 

We know the following (Pearson, 1911): 


$$
\begin{aligned}
\eta_{p}^2 &= \eta^2
\end{aligned}
$$

Note, that only in the case of the simple design (e.g., one-way ANOVA) the following formula is true. In our case, we are interested in the transformation of the effect size for the interaction between condition A X B. The formula for the eta squared is (e.g., Cohen, 1973, 1988; Fisher, 1925, 1973):

$$
\begin{aligned}
\eta^2 &= \frac{\mathrm{SS}_\mathrm{effect}}{\mathrm{SS}_\mathrm{total}}\\
\end{aligned}
$$

Alternatively, Kennedy (1970) proposed:

$$
\begin{aligned}
\eta^2 &= \frac{\mathrm{n}_\mathrm{1}F}{\mathrm{n}_\mathrm{1}F + \mathrm{n}_\mathrm{2}}\\
\end{aligned}
$$

which could be simplified to: 

$$
\begin{aligned}
\eta^2 &= \frac{\mathrm{SS}_\mathrm{effect}}{\mathrm{SS}_\mathrm{effect}+\mathrm{SS}_\mathrm{error}}\\  
\end{aligned}
$$

However, Cohen (1973) pointed out that this formula represents partial eta squared, not the actual eta squared. 

Taken together, we do not know the way how to convert the eta-squared to partial eta-squared given the information usually provided in the paper. 

We now are going to consider an alternative approach.

## Converting partial eta-squared to Cohen's d

Let's take the Cohen's d from the following formula:

$$
\begin{aligned}
\eta_{p}^2 &= \frac{\ d^2 * N}{\ d^2 * N + N\mathrm{-1}}\\
\end{aligned}
$$

After transformations (see more details here: https://haiyangjin.github.io/2020/05/eta2d/) we can get Cohen's d.

$$
\begin{aligned}
\ d = \sqrt[](\frac{(\mathrm{N - 1})}{\mathrm{N}} * \frac{\eta_{p}^2}{(1-\eta_{p}^2)})\\
\end{aligned}
$$

We now consider a set of values that we got from our ANOVA analyses for the simulated dataset. 

```{r}
nice(ANOVAresults, es = 'pes', sig_symbols = rep("", 4))
CohenD <- sqrt((nSubj-1)/nSubj*.407/(1-.407))
CohenD 
```

Cohen's d is equal to 0.8074798. Let's check if this calculation is correct. 

$$
\begin{aligned}
\ t &= \ d * \sqrt[]{N} \\
\end{aligned}
$$
```{r}
tvalue <- CohenD *sqrt(nSubj)
tvalue
```

$$
\begin{aligned}
\ F &= \ t^2\\
\end{aligned}
$$
```{r}
Fvalue <- tvalue*tvalue
Fvalue
nice(ANOVAresults, es = 'pes', sig_symbols = rep("", 4))
```

The F-value is similar to the F-value presented in the ANOVA table!

## Converting Cohen's d to Hedges' g. 

To transform Cohen's d to Hedges' g, we used mentioned above 'esc' package. 

```{r echo=FALSE}
library('esc')
hedges_g(CohenD, totaln = 20)
```


# Example

In the part-whole task, Tanaka et al. (2004) provided information about significant two-way interaction between Test Type (2: part vs whole) and Race of Target Face (Caucasian faces vs Asian faces) on accuracy for Caucasian (n = 21, F(1,20) = 18.436, MS = .056, p <.001) but not Asian (n = 21, F(1,20) = .073, MS = .001, p >.10) participants.

```{r}
nSubjC <- 21 # Number of Caucasian participants
nSubjA <- 21 # Number of Asian participants

# Partial eta-squared 
n2pA <- .073*1/(.073*1+20)
n2pC <- 18.436*1/(18.436*1+20) 

CohenD_A <- sqrt((nSubjA -1)/nSubjA *n2pA/(1-n2pA))
CohenD_A
gA <- hedges_g(CohenD_A, totaln = nSubjA)
gA

CohenD_C <- sqrt((nSubjC-1)/nSubjC*n2pC/(1-n2pC))
CohenD_C
gC<- hedges_g(CohenD_C, totaln = nSubjC)
gC
```

The Hedges' g for the interaction between Test Type and Race of Target Face was 0.057 and 0.899 for Asian and Caucasian participants, respectively. 


# Some words at the end

We cannot guarantee that these conversions are applicable in all situations, but at least they work well for cases similar to the simulated data.

Additionally, we really appreciate if you could let us know if there is any error in this post.

# References

Cohen, J. (1973). Eta-squared and partial eta-squared in fixed factor ANOVA designs. Educational and Psychological Measurement, 33(1), 107-112.

Cohen, J. (1988). Statistical power analysis for the behavioral sciences (2nd ed.). Hillsdale, NJ: Erlbaum.

Fisher, R.A. (1925). Statistical methods for research workers. Edinburgh: Oliver & Boyd.

Fisher, R. A. (1973). Statistical methods for research workers (14th ed.). New York: Hafner.

Hedges, L., & Olkin, I. (1985). Statistical Methods for Meta-analysis. San Diego, CA: Academic Press.

Kennedy, J. J. (1970). The eta coefficient in complex ANOVA designs. Educational and Psychological Measurement, 30(4), 885-889.

Lakens, D. (2013). Calculating and reporting effect sizes to facilitate cumulative science: a practical primer for t-tests and ANOVAs. Frontiers in psychology, 4, 863.

Pearson, K. (1911). On a correction needful in the case of the correlation ratio. Biometrika, 8, 254–256.

Tanaka, J. W., Kiefer, M., & Bukach, C. M. (2004). A holistic account of the own-race effect in face recognition: Evidence from a cross-cultural study. Cognition, 93(1), B1-B9.'


`citation('afex')`

`citation('esc')`

`citation('tidyverse')`


